---
title: "Cricket Data Analysis - Batsmen Performance"
author: "Team CricEval"
date: "`r Sys.Date()`"
output: html_document
---

```r
```{r}
library(readr)

# Read dataset
df <- read_csv("top10_batsmen_dataset.csv")

# View first rows
head(df)

# Dimensions (rows, columns)
dim(df)

# Structure
str(df)

# Summary stats
summary(df)

# Column names
colnames(df)




# 1. Missing values per column
missing_values <- colSums(is.na(df))
missing_values

# 2. Empty strings in character columns
empty_strings <- sapply(df, function(x) {
  if (is.character(x)) sum(x == "", na.rm = TRUE) else 0
})
empty_strings

# 3. Duplicate rows
duplicate_count <- sum(duplicated(df))
duplicate_count

# 4. Data types of all columns
data_types <- sapply(df, class)
data_types

# 5. Detect possible mixed/wrong types (numeric stored as character)
mixed_type_check <- sapply(df, function(x) {
  if (is.character(x)) {
    suppressWarnings(sum(!is.na(as.numeric(x))))
  } else {
    NA
  }
})
mixed_type_check
# 🧹 3. Data Cleaning
```{r}
library(dplyr)
library(stringr)

# 1. Remove duplicate rows
df <- df %>%
  distinct()

# 2. Trim whitespace from character columns
df <- df %>%
  mutate(across(where(is.character), str_trim))

# 3. Standardize text casing (lowercase for consistency)
df <- df %>%
  mutate(across(where(is.character), str_to_lower))

# 4. Convert numeric-like columns to numeric (EDIT based on your dataset)
# Example:
# df$runs_off_bat <- as.numeric(df$runs_off_bat)
# df$extras <- as.numeric(df$extras)

# 5. Convert categorical columns to factor (EDIT based on dataset)
# Example:
# df$batsman <- as.factor(df$batsman)
# df$bowler <- as.factor(df$bowler)
# df$venue <- as.factor(df$venue)

# 6. Handle invalid values (example rules)
# Replace negative runs with NA (if any)
# df$runs_off_bat[df$runs_off_bat < 0] <- NA

# Replace empty strings with NA
df[df == ""] <- NA
```{r}
# Remove rows where BOTH ball_line and ball_length are missing
df <- df %>%
  filter(!(is.na(ball_line) & is.na(ball_length)))

# Check how many rows remain
dim(df)
# 7. Final check
str(df)
summary(df)
# 🗑️ 4. Drop or Ignore Useless Columns
```{r}
library(dplyr)

# Drop irrelevant / redundant columns
df <- df %>%
  select(
    -`full name_striker`,
    -`full name_bowler`,
    -`image url_striker`,
    -`image url_bowler`,
    -`major teams_striker`,
    -`major teams_bowler`,
    -`player_of_match.1`,
    -event,
    -type
  )

# Optional: rename columns with spaces for easier use later
df <- df %>%
  rename(
    balls_faced = `striker balls faced`,
    total_runs = `total striker runs`,
    out_runs = `player out runs`,
    out_balls = `player out balls faced`
  )

# Verify updated columns
colnames(df)

# 🏏 5. Understand the Structure of the Cricket Data

```{r}
library(dplyr)

# 1. Unique batsmen
num_batsmen <- n_distinct(df$striker)
num_batsmen

# 2. Matches, seasons, venues, opponents
num_matches <- n_distinct(df$match_id)
num_seasons <- n_distinct(df$season)
num_venues <- n_distinct(df$venue)
num_opponents <- n_distinct(df$bowling_team)

num_matches
num_seasons
num_venues
num_opponents

# 3. Understand dataset level (by checking uniqueness)
# Check how many balls per match
df %>%
  group_by(match_id) %>%
  summarise(total_balls = n()) %>%
  head()

# Check balls per batsman
df %>%
  group_by(striker) %>%
  summarise(balls_faced = n()) %>%
  arrange(desc(balls_faced)) %>%
  head()

# 4. Identify column types (manual understanding)
colnames(df)

# 5. Decide analysis level

# Ball-level data example
head(df)

# Batsman-level aggregation
batsman_summary <- df %>%
  group_by(striker) %>%
  summarise(
    total_runs = sum(runs_off_bat, na.rm = TRUE),
    balls = n(),
    strike_rate = (total_runs / balls) * 100
  )

head(batsman_summary)

# Match-level summary
match_summary <- df %>%
  group_by(match_id) %>%
  summarise(
    total_runs = sum(runs_off_bat, na.rm = TRUE),
    total_wickets = sum(wicket, na.rm = TRUE)
  )

head(match_summary)

# 📊 7. Univariate EDA

```{r}
library(dplyr)
library(ggplot2)

# 1. Batsman frequency
df %>%
  count(striker, sort = TRUE) %>%
  head(10)

# 2. Bowler frequency
df %>%
  count(bowler, sort = TRUE) %>%
  head(10)

# 3. Venue frequency
df %>%
  count(venue, sort = TRUE) %>%
  head(10)

# 4. Season frequency
df %>%
  count(season, sort = TRUE)

# 5. Batting team frequency
df %>%
  count(batting_team, sort = TRUE)

# 6. Bowling team frequency
df %>%
  count(bowling_team, sort = TRUE)

# 7. Shot type frequency
df %>%
  count(shot_played, sort = TRUE) %>%
  head(10)

# 8. Shot direction frequency
df %>%
  count(shot_direction, sort = TRUE) %>%
  head(10)

# 9. Wicket type frequency
df %>%
  count(wicket_type, sort = TRUE)

```{r}
library(ggplot2)

# 1. Runs scored per ball
ggplot(df, aes(x = runs_off_bat)) +
  geom_histogram(bins = 10) +
  labs(title = "Distribution of Runs per Ball")

# 2. Extras distribution
ggplot(df, aes(x = extras)) +
  geom_histogram(bins = 10) +
  labs(title = "Distribution of Extras")

# 3. Balls faced (using aggregated data)
batsman_stats <- df %>%
  group_by(striker) %>%
  summarise(
    balls = n(),
    total_runs = sum(runs_off_bat, na.rm = TRUE),
    strike_rate = (total_runs / balls) * 100
  )

# Balls faced distribution
ggplot(batsman_stats, aes(x = balls)) +
  geom_histogram(bins = 20) +
  labs(title = "Distribution of Balls Faced")

# 4. Total runs distribution
ggplot(batsman_stats, aes(x = total_runs)) +
  geom_histogram(bins = 20) +
  labs(title = "Distribution of Total Runs")

# 5. Strike rate distribution
ggplot(batsman_stats, aes(x = strike_rate)) +
  geom_histogram(bins = 20) +
  labs(title = "Distribution of Strike Rate")

# 🔗 8. Bivariate EDA

```{r}
library(dplyr)
library(ggplot2)

# Create batsman summary (reuse this everywhere)
batsman_stats <- df %>%
  group_by(striker) %>%
  summarise(
    total_runs = sum(runs_off_bat, na.rm = TRUE),
    balls = n(),
    strike_rate = (total_runs / balls) * 100
  ) %>%
  filter(balls > 50)   # remove very small samples

# 1. Compare runs by batsman (Top 10)
batsman_stats %>%
  arrange(desc(total_runs)) %>%
  head(10) %>%
  ggplot(aes(x = reorder(striker, total_runs), y = total_runs)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top 10 Batsmen by Total Runs")

# 2. Compare strike rate by batsman
batsman_stats %>%
  arrange(desc(strike_rate)) %>%
  head(10) %>%
  ggplot(aes(x = reorder(striker, strike_rate), y = strike_rate)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top 10 Batsmen by Strike Rate")

# 3. Performance by venue
venue_stats <- df %>%
  group_by(venue) %>%
  summarise(avg_runs = mean(runs_off_bat, na.rm = TRUE))

ggplot(venue_stats, aes(x = reorder(venue, avg_runs), y = avg_runs)) +
  geom_col() +
  coord_flip() +
  labs(title = "Average Runs per Ball by Venue")

# 4. Performance by opponent (bowling team)
opponent_stats <- df %>%
  group_by(bowling_team) %>%
  summarise(avg_runs = mean(runs_off_bat, na.rm = TRUE))

ggplot(opponent_stats, aes(x = reorder(bowling_team, avg_runs), y = avg_runs)) +
  geom_col() +
  coord_flip() +
  labs(title = "Performance vs Opponent")

# 5. Performance by season
season_stats <- df %>%
  group_by(season) %>%
  summarise(avg_runs = mean(runs_off_bat, na.rm = TRUE))

ggplot(season_stats, aes(x = season, y = avg_runs)) +
  geom_line() +
  geom_point() +
  labs(title = "Performance Across Seasons")

# 6. Performance by batting style
bat_style_stats <- df %>%
  group_by(`batting style_striker`) %>%
  summarise(avg_runs = mean(runs_off_bat, na.rm = TRUE))

ggplot(bat_style_stats, aes(x = `batting style_striker`, y = avg_runs)) +
  geom_col() +
  labs(title = "Performance by Batting Style")

# 7. Performance by bowling style
bowl_style_stats <- df %>%
  group_by(`bowling style_bowler`) %>%
  summarise(avg_runs = mean(runs_off_bat, na.rm = TRUE))

ggplot(bowl_style_stats, aes(x = `bowling style_bowler`, y = avg_runs)) +
  geom_col() +
  labs(title = "Performance vs Bowling Style")

# 8. Shot type vs runs
shot_stats <- df %>%
  group_by(shot_played) %>%
  summarise(avg_runs = mean(runs_off_bat, na.rm = TRUE))

ggplot(shot_stats, aes(x = reorder(shot_played, avg_runs), y = avg_runs)) +
  geom_col() +
  coord_flip() +
  labs(title = "Runs by Shot Type")

# 9. Shot direction vs runs
direction_stats <- df %>%
  group_by(shot_direction) %>%
  summarise(avg_runs = mean(runs_off_bat, na.rm = TRUE))

ggplot(direction_stats, aes(x = reorder(shot_direction, avg_runs), y = avg_runs)) +
  geom_col() +
  coord_flip() +
  labs(title = "Runs by Shot Direction")

# 📊 9. Visualizations
```{r}
library(dplyr)
library(ggplot2)
library(GGally)

# Prepare batsman stats (reuse)
batsman_stats <- df %>%
  group_by(striker) %>%
  summarise(
    total_runs = sum(runs_off_bat, na.rm = TRUE),
    balls = n(),
    strike_rate = (total_runs / balls) * 100
  ) %>%
  filter(balls > 50)

# -------------------------------
# 1. Bar Chart (Category Counts)
# -------------------------------
df %>%
  count(batting_team) %>%
  ggplot(aes(x = reorder(batting_team, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(title = "Matches by Batting Team")

# -------------------------------
# 2. Histogram (Numeric)
# -------------------------------
ggplot(df, aes(x = runs_off_bat)) +
  geom_histogram(bins = 10) +
  labs(title = "Runs Distribution")

# -------------------------------
# 3. Boxplot (Numeric)
# -------------------------------
ggplot(df, aes(y = runs_off_bat)) +
  geom_boxplot() +
  labs(title = "Boxplot of Runs")

# -------------------------------
# 4. Violin Plot (Batsman Performance)
# -------------------------------
top_batsmen <- batsman_stats %>%
  arrange(desc(total_runs)) %>%
  head(5) %>%
  pull(striker)

df %>%
  filter(striker %in% top_batsmen) %>%
  ggplot(aes(x = striker, y = runs_off_bat)) +
  geom_violin() +
  labs(title = "Batsman Performance Distribution")

# -------------------------------
# 5. Heatmap (Correlation)
# -------------------------------
numeric_df <- batsman_stats %>%
  select(total_runs, balls, strike_rate)

cor_matrix <- cor(numeric_df)

cor_df <- as.data.frame(as.table(cor_matrix))

ggplot(cor_df, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  labs(title = "Correlation Heatmap")

# -------------------------------
# 6. Scatter Plot
# -------------------------------
ggplot(batsman_stats, aes(x = balls, y = total_runs)) +
  geom_point() +
  labs(title = "Balls vs Runs")

# -------------------------------
# 7. Pair Plot
# -------------------------------
ggpairs(numeric_df)

# -------------------------------
# 8. Stacked Bar Chart
# -------------------------------
df %>%
  count(batting_team, shot_played) %>%
  ggplot(aes(x = batting_team, y = n, fill = shot_played)) +
  geom_bar(stat = "identity") +
  labs(title = "Shot Type by Team") +
  coord_flip()

# ❗ 10. Missing Value and Null Analysis

```{r}
library(dplyr)
library(ggplot2)

# 1. Count missing values per column
missing_values <- colSums(is.na(df))
missing_values

# 2. Percentage of missing values
missing_percent <- (missing_values / nrow(df)) * 100
missing_percent

# 3. Visualize missing values
missing_df <- data.frame(
  column = names(missing_values),
  missing = missing_values
)

ggplot(missing_df, aes(x = reorder(column, missing), y = missing)) +
  geom_col() +
  coord_flip() +
  labs(title = "Missing Values per Column")

# 4. Handle missing values

# Replace NA in categorical columns with "unknown"
df <- df %>%
  mutate(across(where(is.character), ~ifelse(is.na(.), "unknown", .)))

# Replace NA in numeric columns with median
df <- df %>%
  mutate(across(where(is.numeric), ~ifelse(is.na(.), median(., na.rm = TRUE), .)))

# 5. Final check
colSums(is.na(df))