---
title: "Cricket Data Analysis - Batsman Performance"
author: "Team CricEval"
date: "`r Sys.Date()`"
output: html_document
---

# Cricket Data Analysis - Batsman Performance

### Installing and Loading Packages

```{r}
#install.packages(c(
#  "tidyverse",
#  "data.table",
#  "ggplot2",
#  "corrplot",
#  "caret",
#  "cluster",
#  "GGally",
#  "reshape2",
#  "xgboost",
#  "Matrix",
#  "janitor",
#))
```

```{r}
library(tidyverse)
library(data.table)
library(ggplot2)
library(corrplot)
library(caret)
library(cluster)
library(GGally)
library(reshape2)
library(stringr)
library(xgboost)
library(Matrix)
library(janitor)
```

### Load and Display Dataset

```{r}
df <- read_csv("top10_batsmen_dataset.csv")
head(df)
dim(df)
str(df)
```

### Summarize the Dataset

```{r}
summary(df)
```

### Understand the Structure of the Cricket Data

```{r}
print(tibble(column_names = names(df)), n = Inf)
```

```{r}
# Understanding all unique values of all features
df %>% distinct(season)
df %>% distinct(venue)
df %>% distinct(striker)
df %>% distinct(non_striker)
df %>% distinct(bowler)
df %>% distinct(shot_direction)
df %>% distinct(wicket_type)
df %>% distinct(other_wicket_type)
df %>% distinct(other_player_dismissed)
df %>% distinct(gender)
df %>% distinct(event)
df %>% distinct(player_of_match)
df %>% distinct(player_of_match.1)
df %>% distinct(outcome)
df %>% distinct(format)
df %>% distinct(type)
df %>% distinct(`batting style_striker`)
df %>% distinct(`bowling style_striker`)
df %>% distinct(`batting style_bowler`)
df %>% distinct(`bowling style_bowler`)
df %>% distinct(batting_team)
df %>% distinct(bowling_team)
df %>% distinct(country_striker)
df %>% distinct(country_bowler)
df %>% distinct(match_number)
df %>% distinct(match_id)
```

### Data Cleaning

```{r}
# Remove unnecessary columns
unwanted_cols <- c(
  "match_number",
  "other_wicket_type",
  "other_player_dismissed",
  "full name_striker",
  "image url_striker",
  "full name_bowler",
  "image url_bowler"
)

df <- df %>% select(-all_of(unwanted_cols))
```

```{r}
# Remove duplicate rows
df[duplicated(df), ]
df <- df %>% distinct() # No duplicates present
```

```{r}
# Trim whitespace from character columns
df <- df %>% mutate(across(where(is.character), str_trim))
```

```{r}
clean_cat <- function(x) {
  x %>%
    str_squish() %>%
    str_trim() %>%
    str_replace_all("&", "and") %>%
    str_replace_all("[^A-Za-z0-9]+", "_") %>%
    str_replace_all("_+", "_") %>%
    str_replace_all("^_|_$", "")
}

df <- df %>%
  mutate(across(where(is.character), clean_cat))

df <- janitor::clean_names(df)
```

```{r}
# Display all columns
str(df)
```

```{r}
str(df)
```

### Univariate EDA

```{r}

# 1. Batsman frequency
df %>% count(striker, sort = TRUE) %>% head(10)

# 2. Bowler frequency
df %>% count(bowler, sort = TRUE) %>% head(10)

# 3. Venue frequency
df %>% count(venue, sort = TRUE) %>% head(10)

# 4. Season frequency
df %>% count(season, sort = TRUE)

# 5. Batting team frequency
df %>% count(batting_team, sort = TRUE)

# 6. Bowling team frequency
df %>% count(bowling_team, sort = TRUE)

# 7. Shot type frequency
df %>% count(shot_played, sort = TRUE) %>% head(10)

# 8. Shot direction frequency
df %>% count(shot_direction, sort = TRUE) %>% head(10)

# 9. Wicket type frequency
df %>% count(wicket_type, sort = TRUE)
```

### Bivariate EDA

```{r}

# Create batsman summary (reuse this everywhere)
batsman_stats <- df %>% 
  group_by(striker) %>% 
  summarise( 
    total_runs = sum(runs_off_bat, na.rm = TRUE), 
    balls = n(), 
    strike_rate = (total_runs / balls) * 100
  ) %>% 
  filter(balls > 50) # remove very small samples

# 1. Compare runs by batsman (Top 10)
batsman_stats %>% 
  arrange(desc(total_runs)) %>% 
  head(10) %>% 
  ggplot(aes(x = reorder(striker, total_runs), y = total_runs)) + 
  geom_col() + 
  coord_flip() + 
  labs(
    title = "Top 10 Batsmen by Total Runs"
  )

# 2. Compare strike rate by batsman
batsman_stats %>% 
  arrange(desc(strike_rate)) %>% 
  head(10) %>% 
  ggplot(aes(x = reorder(striker, strike_rate), y = strike_rate)) + 
  geom_col() + 
  coord_flip() + 
  labs(
    title = "Top 10 Batsmen by Strike Rate"
  )

# 3. Performance by venue
venue_stats <- df %>% 
  group_by(venue) %>% 
  summarise(avg_runs = mean(runs_off_bat, na.rm = TRUE))

ggplot(venue_stats, aes(x = reorder(venue, avg_runs), y = avg_runs)) + 
  geom_col() + 
  coord_flip() + 
  labs(
    title = "Average Runs per Ball by Venue")

# 4. Performance by opponent (bowling team)
opponent_stats <- df %>%
  group_by(bowling_team) %>% 
  summarise(avg_runs = mean(runs_off_bat, na.rm = TRUE))

ggplot(opponent_stats, aes(x = reorder(bowling_team, avg_runs), y = avg_runs)) +
  geom_col() + 
  coord_flip() + 
  labs(
    title = "Performance vs Opponent"
  )

# 5. Performance by season
season_stats <- df %>% 
  group_by(season) %>% 
  summarise(avg_runs = mean(runs_off_bat, na.rm = TRUE))

ggplot(season_stats, aes(x = season, y = avg_runs)) +
  geom_line() + 
  geom_point() + 
  labs(
    title = "Performance Across Seasons"
  )

# 6. Performance by batting style
bat_style_stats <- df %>% 
  group_by(batting_style_striker) %>% 
  summarise(avg_runs = mean(runs_off_bat, na.rm = TRUE))

ggplot(bat_style_stats, aes(x = batting_style_striker, y = avg_runs)) + 
  geom_col() + 
  labs(
    title = "Performance by Batting Style"
  )

# 7. Performance by bowling style
bowl_style_stats <- df %>% 
  group_by(bowling_style_bowler) %>% 
  summarise(avg_runs = mean(runs_off_bat, na.rm = TRUE))

ggplot(bowl_style_stats, aes(x = bowling_style_bowler, y = avg_runs)) + 
  geom_col() + 
  labs(
    title = "Performance vs Bowling Style"
  )

# 8. Shot type vs runs
shot_stats <- df %>% 
  group_by(shot_played) %>% 
  summarise(avg_runs = mean(runs_off_bat, na.rm = TRUE))

ggplot(shot_stats, aes(x = reorder(shot_played, avg_runs), y = avg_runs)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Runs by Shot Type"
  )

# 9. Shot direction vs runs
direction_stats <- df %>%
  group_by(shot_direction) %>%
  summarise(avg_runs = mean(runs_off_bat, na.rm = TRUE))

ggplot(direction_stats, aes(x = reorder(shot_direction, avg_runs), y = avg_runs)) +
  geom_col() + 
  coord_flip() + 
  labs(
    title = "Runs by Shot Direction"
  )
```

### Visualization

```{r}
## Bar Chart for Category Counts
# Batsman frequency (Stacked bar chart)
ggplot(df, aes(x = striker, fill = batting_team)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(
    title = "Ball Count per Batsman",
    x = "Batsman",
    y = "Ball Count",
    fill = "Batting Team"
  )
# From this plot, we can
# - Find who played the most balls
# - Identify multi-team players
# - Identify role of the batsman
#   - High count -> anchor player
#   - Low count -> finisher / tail-ender

# Batting team
ggplot(df %>% count(batting_team),
       aes(x = reorder(batting_team, n), y = n, fill = batting_team)) +
  geom_bar(stat="identity") +
  coord_flip() +
  labs(
    title = "Balls Faced by Each Team",
    x = "Balls",
    y = "Batting Team",
    fill = "Batting Team"
  )
# From this plot, we can
# - Identify which team appears most in the dataset
# - Identify dataset imbalance

# geom_bar() is used to create bar charts
```

```{r}
# Check how many balls per match
match_stats <- df %>%
  group_by(match_id) %>%
  summarise(total_balls = n())

ggplot(match_stats, aes(x = total_balls)) +
  geom_histogram(bins = 30, fill="blue", color="black") +
  labs(
    title = "Distribution of Total Balls per Match",
    x = "Total Balls",
    y = "Count of Matches"
  )
```

```{r}
# Check balls per batsman
df %>% group_by(striker) %>% 
  summarise(balls_faced = n()) %>% 
  arrange(desc(balls_faced))
```

```{r}
# Batsman-level aggregation
batsman_summary <- df %>% 
  group_by(striker) %>% 
  summarise( 
    total_runs = sum(runs_off_bat, na.rm = TRUE),
    balls = n(),
    strike_rate = (total_runs / balls) * 100
  )

batsman_summary

ggplot(batsman_summary, aes(x = balls, y = total_runs, color = strike_rate)) +
  geom_point(alpha = 0.7) +
  geom_text(aes(label = striker), vjust = -0.5, size = 3) +
  labs(
    title = "Batsman Performance",
    x = "Total Balls Played",
    y = "Total Runs"
  ) +
  theme_minimal()
```

```{r}
# Match-level summary
match_summary <- df %>% 
  group_by(match_id) %>% 
  summarise(
    total_runs = sum(runs_off_bat, na.rm = TRUE),
    total_wickets = sum(wicket, na.rm = TRUE)
)
head(match_summary)

ggplot(match_summary, aes(x = total_wickets, y = total_runs)) +
  geom_point(alpha = 0.6, color = "blue") +
  labs(
    title = "Match Analysis: Runs vs Wickets",
    x = "Total Wickets",
    y = "Total Runs"
  ) +
  theme_minimal()
```

```{r}
# Runs distribution
ggplot(df, aes(x = runs_off_bat)) +
  geom_histogram(bins = 20, fill='blue') +
  labs(
    title = "Distribution of Runs per Ball",
    x = "Runs off bat",
    y = "Count"
  )

# Extras distribution
ggplot(df, aes(x = extras)) +
  geom_histogram(bins = 20, fill="green") +
  labs(
    title = "Distribution of Extras",
    x = "Extra",
    y = "Count"
  )
```

```{r}
ggplot(df, aes(x = striker, y = runs_off_bat)) +
  geom_violin(fill = "lightblue") +
  theme(axis.text.x = element_text(angle = 90)) +
  ggtitle("Batsman Performance")
# Width of violin
# - Wider → more frequent values
# - Narrow → rare values
# - Example
#   - Wide at 0 → many dot balls
#   - Wide at 4 → many boundaries
```

```{r}
# Runs vs extras
ggplot(df, aes(x = extras, y = runs_off_bat)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Runs vs Extra per Ball",
    x = "Extra run team got",
    y = "Runs batsman scored on that ball"
  )
# This helps us learn about the different types of extras
# - No balls (with six, with four, dot ball, etc)
# - Wide (...)
# - byes (...)

# Balls faced vs runs (after aggregation)
batsman_df <- df %>%
  group_by(striker) %>%
  summarise(
    total_runs = sum(runs_off_bat),
    balls = n()
  )
```

```{r}
# Check missing values
colSums(is.na(df))

# Notice that all NA values are expected to be there
```

```{r}
# Outlier Detection
batsman_df <- df %>%
  group_by(striker) %>%
  summarise(
    total_runs = sum(runs_off_bat),
    balls = n(),
    strike_rate = (total_runs / balls) * 100
  )
```

```{r}
detect_outliers <- function(x) {
  Q1 <- quantile(x, 0.25)
  Q3 <- quantile(x, 0.75)
  IQR <- Q3 - Q1
  
  lower <- Q1 - 1.5 * IQR
  upper <- Q3 + 1.5 * IQR
  
  return(x < lower | x > upper)
}

batsman_df <- batsman_df %>%
  mutate(
    outlier_sr = detect_outliers(strike_rate),
    outlier_runs = detect_outliers(total_runs),
    outlier_balls = detect_outliers(balls)
  )

outliers_df <- batsman_df %>%
  filter(outlier_sr | outlier_runs | outlier_balls)

outliers_df
```

```{r}
# runs_off_bat
ggplot(df, aes(y = runs_off_bat)) +
  geom_boxplot(fill = "skyblue") +
  ggtitle("Runs per Ball per Batsman")

# strike rate
ggplot(batsman_df, aes(y = strike_rate)) +
  geom_boxplot(fill = "lightgreen") +
  ggtitle("Strike Rate per Batsman")

# balls faced
ggplot(batsman_df, aes(y = balls)) +
  geom_boxplot(fill = "orange") +
  ggtitle("Balls Faced per Batsman")

# total runs
ggplot(batsman_df, aes(y = total_runs)) +
  geom_boxplot(fill = "pink") +
  ggtitle("Total Runs per Batsman")
```

### Model Training

#### 1. Shot to Play - XGBoost

**Why make striker stats and bowler stats?**

If we directly use the striker and bowler names (even if one hot encoded), the model will treat them as states and will overfit to the names. We do not want that, so we will rather train the model on their stats and playing styles.

```{r}
df <- df %>%
  mutate(
    is_boundary = ifelse(runs_off_bat %in% c(4,6), 1, 0),
    total_runs_conceded = runs_off_bat + extras,
    is_dot = ifelse(runs_off_bat == 0, 1, 0)
  )
```

```{r}
striker_stats <- df %>%
  group_by(striker) %>%
  summarise(
    total_runs = sum(runs_off_bat, na.rm = TRUE),
    balls_faced = n(),
    strike_rate = (total_runs / balls_faced) * 100,
    boundary_pct = mean(is_boundary, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
bowler_stats <- df %>%
  group_by(bowler) %>%
  summarise(
    runs_conceded = sum(total_runs_conceded, na.rm = TRUE),
    balls_bowled = n(),
    wickets = sum(wicket, na.rm = TRUE),
    economy = runs_conceded / (balls_bowled / 6),
    wicket_rate = wickets / balls_bowled,
    dot_ball_pct = mean(is_dot, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
df <- df %>%
  left_join(striker_stats, by = "striker") %>%
  left_join(bowler_stats, by = "bowler")
```

```{r}
train_dataset <- df %>%
  filter(!is.na(shot_played)) %>%
  select(
    shot_played,
    ball_length,
    ball_line,
    innings,
    ball,
    strike_rate,
    boundary_pct,
    economy,
    wicket_rate,
    dot_ball_pct,
    batting_style_striker,
    bowling_style_bowler
  ) %>%
  na.omit()

train_dataset
```

```{r}
train_dataset %>% 
  distinct(shot_played)
```

**Label Encoding**

```{r}
# Encode Target (shot_played)
encode_target <- function(df, target_col) {
  df[[target_col]] <- as.factor(df[[target_col]])
  label <- as.numeric(df[[target_col]]) - 1
  
  return(list(
    df = df,
    label = label
  ))
}

result <- encode_target(train_dataset, "shot_played")

train_dataset <- result$df
label <- result$label

tibble(label)
```

**One Hot Encoding**

We will one hot encode **categorical features** as model only understands numerical data.

```{r}

onehot_encode_features <- function(df, ignore_col) {
  dummies <- dummyVars(~ ., data = df %>% select(-all_of(ignore_col)))
  X <- predict(dummies, newdata = df) 
  
  return(list(
    X = as.matrix(X),
    encoder = dummies
  ))
}

result <- onehot_encode_features(train_dataset, "shot_played")

X <- result$X
encoder <- result$encoder

tibble(colnames(df))
X
```

```{r}
train_test_split <- function(X, y, train_ratio = 0.8) {
  set.seed(42)
  
  train_idx <- sample(seq_len(nrow(X)), size = train_ratio * nrow(X))
  
  X_train <- X[train_idx, ]
  X_test  <- X[-train_idx, ]
  
  y_train <- y[train_idx]
  y_test  <- y[-train_idx]
  
  return(list(
    X_train = X_train,
    X_test = X_test,
    y_train = y_train,
    y_test = y_test
  ))
}

result <- train_test_split(X, label)

X_train <- result$X_train
X_test  <- result$X_test
y_train <- result$y_train
y_test  <- result$y_test
```

```{r}
train_xgboost_model <- function(X_train, y_train, X_test, y_test, 
                                nrounds = 100, max_depth = 6, eta = 0.1) {
  
  # Create DMatrix
  dtrain <- xgb.DMatrix(data = X_train, label = y_train)
  dtest  <- xgb.DMatrix(data = X_test, label = y_test)
  
  # Number of classes
  num_class <- length(unique(y_train))
  
  # Parameters
  params <- list(
    objective = "multi:softprob",   # multiclass classification (probabilities)
    num_class = num_class,          # number of target classes
    eval_metric = "mlogloss",       # multiclass log loss
    max_depth = 6,                  # depth of decision tree
    eta = 0.1                       # learning rate
  )
  # objective = "multi:softmax" -> directly predict the most probable class
  # This is not so flexibly, we wont be able to take topN classes
    
  # Train model
  model <- xgb.train(
    params = params,
    data = dtrain,
    nrounds = nrounds,
    evals = list(train = dtrain, test = dtest),
    verbose = 1
  )
  
  eval_log <- attr(model, "evaluation_log")
  
  return(list(
    model = model,
    eval_log = eval_log,
    params = params,
    dtrain = dtrain,
    dtest = dtest,
    num_class = num_class
  ))
}
```

```{r}
result <- train_xgboost_model(X_train, y_train, X_test, y_test)

model = result$model
eval_log = result$eval_log
params = result$params
dtest = result$dtest
```

```{r}
str(model)
```

```{r}
str(eval_log)
```

```{r}
plot_model_loss <- function(eval_log) {
  log_long <- eval_log %>%
    pivot_longer(
      cols = c(train_mlogloss, test_mlogloss),
      names_to = "dataset",
      values_to = "loss"
    )
  
  ggplot(log_long, aes(x = iter, y = loss, color = dataset)) +
    geom_line(linewidth = 1) +
    labs(
      title = "Train vs Test Loss",
      x = "Iteration",
      y = "Log Loss",
      color = "Dataset"
    ) +
    theme_minimal()
}
```

```{r}
plot_model_loss(eval_log)
```

```{r}
calc_accuracy <- function(model, dtest, y_test, num_class) {
  pred_probs <- predict(model, dtest)
  
  pred_matrix <- matrix(pred_probs, ncol = num_class, byrow = TRUE)
  pred_labels <- max.col(pred_matrix) - 1
  
  accuracy <- mean(pred_labels == y_test)
  
  return(list(
    pred_labels = pred_labels,
    accuracy = accuracy
  ))
}
```

```{r}
num_class <- length(unique(y_test))

result <- calc_accuracy(model, dtest, y_test, num_class)

accuracy <- result$accuracy
print(accuracy)

# Why accuracy so low?
# -> High number of classes (39 classes)
# -> Reduce number of shot played classes by generalizing some classes together
```

```{r}
importance <- xgb.importance(model = model)
xgb.plot.importance(importance)
```

#### 2. Shot to Play - XGBoost with Grouped Target Classes

```{r}
colnames(train_dataset)
```

```{r}
train_dataset <- train_dataset %>%
  mutate(
    shot_group = case_when(
      shot_played %in% c("defense", "leave") ~ "defensive",

      shot_played %in% c(
        "drive", "cover_drive", "straight_drive", "off_drive", "on_drive"
      ) ~ "drives",

      shot_played %in% c(
        "cut", "late_cut", "square_drive", "upper_cut",
        "pull", "hook", "sweep", "french_cut"
      ) ~ "backfoot_attack",

      shot_played %in% c(
        "dab", "nudge", "work", "flick", "whips", "clips",
        "leg_glance", "glance", "glide", "paddle", "punch"
      ) ~ "placement",

      shot_played %in% c(
        "slog", "loft", "lofted_drive", "ramp", "scoop", "switch_hit"
      ) ~ "aggressive_lofted",

      shot_played %in% c(
        "outside_edge", "inside_edge", "edge", "edged", "beated", "beaten"
      ) ~ "edges_misses",

      TRUE ~ "other"
    )
  )

train_dataset <- train_dataset %>%
  select(-shot_played)

train_dataset %>%
  distinct(shot_group)
```

```{r}
# Sorted Bar Chart
train_dataset %>%
  count(shot_group) %>%
  ggplot(aes(x = reorder(shot_group, n), y = n, fill = shot_group)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Distribution of Shot Groups",
    x = "Shot Group",
    y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Donut chart
train_dataset %>%
  count(shot_group) %>%
  mutate(percent = n / sum(n) * 100) %>%
  ggplot(aes(x = 2, y = percent, fill = shot_group)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  xlim(0.5, 2.5) +
  theme_void() +
  labs(title = "Shot Group Distribution")
```

```{r}
result <- encode_target(train_dataset, "shot_group")

train_dataset <- result$df
label <- result$label

tibble(label)
```

```{r}
result <- onehot_encode_features(train_dataset, "shot_group")

X <- result$X
encoder <- result$encoder

tibble(colnames(df))
X
```

```{r}
result <- train_test_split(X, label)

X_train <- result$X_train
X_test  <- result$X_test
y_train <- result$y_train
y_test  <- result$y_test
```

```{r}
result <- train_xgboost_model(X_train, y_train, X_test, y_test)

model = result$model
eval_log = result$eval_log
params = result$params
dtest = result$dtest
```

```{r}
plot_model_loss(eval_log)
```

```{r}
num_class <- length(unique(y_test))

result <- calc_accuracy(model, dtest, y_test, num_class)

accuracy <- result$accuracy
print(accuracy)
```

#### 3. Will batsman score a run or not - Binary Classification

```{r}
colnames(df)
```

```{r}
df$target <- df$is_dot

model_df <- df %>%
  select(
    target,                
    season,
    venue,
    innings,
    ball,
    batting_team,
    bowling_team,
    ball_length,
    ball_line,
    gender,
    event,
    toss_winner,
    toss_decision,
    format,
    type,
    country_striker,
    batting_style_striker,
    bowling_style_striker,
    playing_role_striker,
    major_teams_striker,
    country_bowler,
    batting_style_bowler,
    bowling_style_bowler,
    playing_role_bowler,
    major_teams_bowler
  ) %>%
  drop_na()
```

```{r}
# Label encoding
model_df <- model_df %>%
  mutate(across(where(is.character), as.factor))
```

```{r}
set.seed(42)
idx <- createDataPartition(model_df$target, p = 0.8, list = FALSE)

train_df <- model_df[idx, ]
test_df  <- model_df[-idx, ]
```

```{r}
x_train <- model.matrix(target ~ ., data = train_df)[, -1]
y_train <- train_df$target

x_test  <- model.matrix(target ~ ., data = test_df)[, -1]
y_test  <- test_df$target

# Fix column mismatch
missing_cols <- setdiff(colnames(x_train), colnames(x_test))
for (col in missing_cols) {
  x_test <- cbind(x_test, rep(0, nrow(x_test)))
  colnames(x_test)[ncol(x_test)] <- col
}
x_test <- x_test[, colnames(x_train)]
```

```{r}
model <- glm(y_train ~ ., data = as.data.frame(x_train), family = "binomial")
```

```{r}
prob <- predict(model, newdata = as.data.frame(x_test), type = "response")
pred <- ifelse(prob > 0.5, 1, 0)
```

```{r}
conf_mat <- table(Predicted = pred, Actual = y_test)
print(conf_mat)
```

```{r}
TP <- conf_mat["1","1"]
TN <- conf_mat["0","0"]
FP <- conf_mat["1","0"]
FN <- conf_mat["0","1"]

accuracy <- (TP + TN) / (TP + TN + FP + FN)
precision <- TP / (TP + FP)
recall <- TP / (TP + FN)
f1 <- 2 * (precision * recall) / (precision + recall)

cat("Accuracy:", accuracy, "\n")
cat("Precision:", precision, "\n")
cat("Recall:", recall, "\n")
cat("F1:", f1, "\n")
```

```{r}
cm_df <- as.data.frame(conf_mat)

ggplot(cm_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), color = "white", size = 6) +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(title = "Confusion Matrix") +
  theme_minimal()
```

```{r}
summary(model)
```

### Statistical Tests

```{r}
df <- df %>%
  mutate(
    shot_group = case_when(
      shot_played %in% c("defense", "leave") ~ "defensive",

      shot_played %in% c(
        "drive", "cover_drive", "straight_drive", "off_drive", "on_drive"
      ) ~ "drives",

      shot_played %in% c(
        "cut", "late_cut", "square_drive", "upper_cut",
        "pull", "hook", "sweep", "french_cut"
      ) ~ "backfoot_attack",

      shot_played %in% c(
        "dab", "nudge", "work", "flick", "whips", "clips",
        "leg_glance", "glance", "glide", "paddle", "punch"
      ) ~ "placement",

      shot_played %in% c(
        "slog", "loft", "lofted_drive", "ramp", "scoop", "switch_hit"
      ) ~ "aggressive_lofted",

      shot_played %in% c(
        "outside_edge", "inside_edge", "edge", "edged", "beated", "beaten"
      ) ~ "edges_misses",

      TRUE ~ "other"
    )
  )
```

```{r}
# -----------------------------
# 1) CHI-SQUARE TESTS
# -----------------------------
# Use categorical columns only
cat_cols <- c(
  "shot_group",           # or shot_played
  "shot_direction",
  "batting_style_striker",
  "bowling_style_bowler",
  "venue",
  "batting_team",
  "bowling_team",
  "format",
  "type",
  "outcome",
  "gender",
  "innings"               # treat as categorical if values are small integers like 1,2
)

# Keep only columns that exist in df
cat_cols <- intersect(cat_cols, names(df))

# Convert to factor
df_cat <- df %>%
  mutate(across(all_of(cat_cols), as.factor))

# Cramer's V helper
cramers_v <- function(tab) {
  chi <- suppressWarnings(chisq.test(tab, correct = FALSE))
  n <- sum(tab)
  r <- nrow(tab)
  k <- ncol(tab)
  as.numeric(sqrt(chi$statistic / (n * min(r - 1, k - 1))))
}

# Run chi-square on all pairs
chi_results <- list()

pairs <- combn(cat_cols, 2, simplify = FALSE)

for (p in pairs) {
  a <- p[1]
  b <- p[2]
  tab <- table(df_cat[[a]], df_cat[[b]])

  # skip degenerate tables
  if (nrow(tab) < 2 || ncol(tab) < 2) next

  chi <- suppressWarnings(chisq.test(tab, correct = FALSE))

  chi_results[[paste(a, b, sep = " vs ")]] <- tibble(
    var1 = a,
    var2 = b,
    chi_sq = as.numeric(chi$statistic),
    df = as.numeric(chi$parameter),
    p_value = chi$p.value,
    cramers_v = cramers_v(tab)
  )
}

chi_results_df <- bind_rows(chi_results) %>%
  arrange(p_value)

print(chi_results_df)

# Example: a single chi-square test
tab <- table(df$shot_group, df$shot_direction)
chisq.test(tab)


# -----------------------------
# 2) CORRELATION TESTS
# -----------------------------
# Use numeric columns only
num_cols <- c(
  "runs_off_bat",
  "extras",
  "ball_length",
  "ball_line",
  "innings",
  "ball",
  "strike_rate",
  "boundary_pct",
  "economy",
  "wicket_rate",
  "dot_ball_pct",
  "total_runs_conceded",
  "is_boundary",
  "is_dot"
)

num_cols <- intersect(num_cols, names(df))

df_num <- df %>%
  select(all_of(num_cols)) %>%
  mutate(across(everything(), as.numeric))

# Spearman is safer for cricket data because many relationships are not perfectly linear
cor_mat <- cor(df_num, use = "pairwise.complete.obs", method = "spearman")
print(cor_mat)

corrplot(cor_mat, method = "color", type = "upper", tl.cex = 0.8)

# Pairwise correlation test example
cor.test(df$runs_off_bat, df$extras, method = "spearman")

# Find strongest correlations
cor_pairs <- as.data.frame(as.table(cor_mat)) %>%
  filter(Var1 != Var2) %>%
  mutate(abs_cor = abs(Freq)) %>%
  arrange(desc(abs_cor)) %>%
  distinct(pmin(as.character(Var1), as.character(Var2)),
           pmax(as.character(Var1), as.character(Var2)), .keep_all = TRUE)

print(head(cor_pairs, 10))
```

```{r}
chisq.test(table(df$shot_group, df$shot_direction))
chisq.test(table(df$shot_group, df$batting_style_striker))
chisq.test(table(df$shot_group, df$bowling_style_bowler))
chisq.test(table(df$shot_group, df$venue))
chisq.test(table(df$batting_team, df$bowling_team))
```

```{r}
cor.test(df$runs_off_bat, df$extras, method = "spearman")
cor.test(df$strike_rate, df$boundary_pct, method = "spearman")
cor.test(df$economy, df$wicket_rate, method = "spearman")
cor.test(df$dot_ball_pct, df$wicket_rate, method = "spearman")
```
